---
title: "QC_Salmo_data"
author: "Sofia Rouot"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    theme: spacelab
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Gene-specific bacterial transcriptome profiling


```{r load packages, message = FALSE, warning = FALSE}
library(semla)
library(openxlsx)
library(readxl)
library(patchwork)
```

# QC: gene-level analyses at the spot level linked to bacterial presence (comment 1 Reviewer 1)

Check object
```{r}
se_filtered <- readRDS("se_filtered.rds")
se_filtered@meta.data
```

```{r}
VlnPlot(se_filtered, features = c("nFeature_Spatial", "nCount_Spatial"), group.by = "sample_id", pt.size = 0)

se_filtered$percent.hb <- PercentageFeatureSet(se_filtered, pattern = "^Hba-|^Hbb-|^Hbs|^Hbq1b|^Hbq1a")
p1 <- VlnPlot(se_filtered, group.by = "sample_id", features = "percent.hb", pt.size = 0.05)
p1

```

## 1. Expression of each bacterial gene at the level of individual cells/spots.

Results:
Fraction of bacterial-positive spots: 4.76 %
0.1319172 average salmo genes per spot
0.1707539 average bacterial UMIs per spot
1.874875 average bacterial UMIs per spot, in Salmo+ spots 

```{r}
salmo_genes <- c("SL1344-2863", "SL1344-3753", "SL1344-2762", "SL1344-0848", "SL1344-2841", "SL1344-4483", "SL1344-2183", "SL1344-2852", "SL1344-1340", "SL1344-1198", "SL1344-1334", "SL1344-1027", "SL1344-3729", "SL1344-1161", "SL1344-1825", "SL1344-4255", "SL1344-1268", "SL1344-4266", "SL1344-1311")

se_filtered@meta.data
```

### Keep BA sample only
```{r}
unique.names <- unique(se_filtered$sample_id)
unique.names

se_filtered.list.nfeat <- lapply(unique.names, function(i){SubsetSTData(se_filtered, expression = (sample_id == i))})
se_filtered.list.nfeat

se_filtered_BA <- se_filtered.list.nfeat[[1]]
unique(se_filtered_BA$sample_id)

```

### Plot each gene spatially
```{r}
se_salmo_only= readRDS (file = "STm_genes_only_obj.rds") #Seurat object where only the Salmonella genes were kept. 

se_salmo_only_log <- se_salmo_only
#Log1p of counts matrix
se_salmo_only_log@assays[["Spatial"]]@layers[["counts"]] <- log1p(se_salmo_only_log@assays[["Spatial"]]@layers[["counts"]])
```

```{r}
cyt <- c("SL1344-2863", "SL1344-3753", "SL1344-2762", "SL1344-0848", "SL1344-2841", "SL1344-4483", "SL1344-2183", "SL1344-2852")
cyt_names <- c("sipC", "uhpT", "iroN", "grxA", "sitA", "yjjZ", "fruB", "prgJ")

vac <- c("SL1344-1340", "SL1344-1198", "SL1344-1334", "SL1344-1027", "SL1344-3729", "SL1344-1161", "SL1344-1825", "SL1344-4255")
vac_names <- c("ssaG", "zinT", "sseC", "pipB", "mgtC", "sifA", "znuA", "phoN")
```

```{r}
#raw
plot_list <- list()
for (gene in cyt) {                     #cytosolic only
  plt_raw <- MapFeatures(se_salmo_only, #se_salmo_only_log or se_salmo_only
                 features = gene, 
                 #max_cutoff = 0.99, 
                 override_plot_dims = F, 
                 section_number = 1,
                 colors = RColorBrewer::brewer.pal(n = 9, name = "Reds"), 
                 label_by = "sample_id", 
                 slot = "counts",
                 pt_size = 0.8, 
                 add_scalebar = TRUE,
                 scalebar_gg = scalebar(x = 1000),
                 scalebar_position = c(0.64, 0.08), 
                 scalebar_height = 0.07,
                 scale_alpha = FALSE) &  
        labs(fill="UMI counts per spot") &
        theme(legend.position = "right", 
        legend.text = element_text(angle = 0),
        legend.margin = margin(t = 50, r = 0, b = 100, l = 0))
  plot_list[[gene]] <- plt_raw

  ggsave(file = paste0("/path/", gene, "_STm_raw_UMI_counts_BA.pdf"), plot = plt_raw, dpi = 300, width = 10, height =8)
}


combined_plot <- wrap_plots(plot_list)
combined_plot
ggsave(file = paste0("/path/all_cytosolic_STm_raw_UMI_counts_BA.pdf"), plot = combined_plot, dpi = 300, width = 30, height =20)

plt_raw
# , image_use = "raw"
```

```{r}
#raw VAC
plot_list <- list()
for (gene in vac) {
  plt_raw <- MapFeatures(se_salmo_only, #se_salmo_only_log or se_salmo_only
                 features = gene, 
                 #max_cutoff = 0.99, 
                 override_plot_dims = F, 
                 section_number = 1,
                 colors = RColorBrewer::brewer.pal(n = 9, name = "Reds"), 
                 label_by = "sample_id", 
                 slot = "counts",
                 pt_size = 0.8, 
                 add_scalebar = TRUE,
                 scalebar_gg = scalebar(x = 1000),
                 scalebar_position = c(0.64, 0.08), 
                 scalebar_height = 0.07,
                 scale_alpha = FALSE) &  
        labs(fill="UMI counts per spot") &
        theme(legend.position = "right", 
        legend.text = element_text(angle = 0),
        legend.margin = margin(t = 50, r = 0, b = 100, l = 0))
  plot_list[[gene]] <- plt_raw

  ggsave(file = paste0("/path/vacuolar/raw/", gene, "_STm_raw_UMI_counts_BA.pdf"), plot = plt_raw, dpi = 300, width = 10, height =8)
}


combined_plot <- wrap_plots(plot_list)
combined_plot
ggsave(file = paste0("/path/vacuolar/raw/all_cytosolic_STm_raw_UMI_counts_BA.pdf"), plot = combined_plot, dpi = 300, width = 30, height =20)

```

```{r}
housekeeping <- c("SL1344-1268", "SL1344-4266", "SL1344-1311")

plot_list <- list()
for (gene in housekeeping) {
  plt_raw <- MapFeatures(se_salmo_only, #se_salmo_only_log or se_salmo_only
                 features = gene, 
                 #max_cutoff = 0.99, 
                 override_plot_dims = F, 
                 section_number = 1,
                 colors = RColorBrewer::brewer.pal(n = 9, name = "Reds"), 
                 label_by = "sample_id", 
                 slot = "counts",
                 pt_size = 0.8, 
                 add_scalebar = TRUE,
                 scalebar_gg = scalebar(x = 1000),
                 scalebar_position = c(0.64, 0.08), 
                 scalebar_height = 0.07,
                 scale_alpha = FALSE) &  
        labs(fill="UMI counts per spot") &
        theme(legend.position = "right", 
        legend.text = element_text(angle = 0),
        legend.margin = margin(t = 50, r = 0, b = 100, l = 0))
  plot_list[[gene]] <- plt_raw

  ggsave(file = paste0("/path/housekeeping/raw/", gene, "_STm_raw_UMI_counts_BA.pdf"), plot = plt_raw, dpi = 300, width = 10, height =8)
}


combined_plot <- wrap_plots(plot_list)
combined_plot
ggsave(file = paste0("/path/raw/housekeeping_STm_raw_UMI_counts_BA.pdf"), plot = combined_plot, dpi = 300, width = 30, height =10)
```


```{r}
#log
plot_list <- list()
for (gene in cyt) {
  plt <- MapFeatures(se_salmo_only_log, #se_salmo_only_log or se_salmo_only
                 features = gene, 
                 #max_cutoff = 0.99, 
                 override_plot_dims = F, 
                 section_number = 1,
                 colors = RColorBrewer::brewer.pal(n = 9, name = "Reds"), 
                 label_by = "sample_id", 
                 slot = "counts",
                 pt_size = 0.8, 
                 add_scalebar = TRUE,
                 scalebar_gg = scalebar(x = 1000),
                 scalebar_position = c(0.64, 0.08), 
                 scalebar_height = 0.07,
                 scale_alpha = FALSE) &  
        labs(fill="log UMI count per spot") &
        theme(legend.position = "right", 
        legend.text = element_text(angle = 0),
        legend.margin = margin(t = 50, r = 0, b = 100, l = 0))
  plot_list[[gene]] <- plt
  ggsave(file = paste0("/path/cytosolic/log/", gene, "_STm_log_UMI_counts_BA.pdf"), plot = plt, dpi = 300, width = 10, height =8)
  
}

combined_plot <- wrap_plots(plot_list)
combined_plot
ggsave(file = paste0("/path/cytosolic/log/all_cytosolic_STm_log_UMI_counts_BA.pdf"), plot = combined_plot, dpi = 300, width = 30, height =20)
```


```{r}
#log VAC
plot_list <- list()
for (gene in vac) {
  plt <- MapFeatures(se_salmo_only_log, #se_salmo_only_log or se_salmo_only
                 features = gene, 
                 #max_cutoff = 0.99, 
                 override_plot_dims = F, 
                 section_number = 1,
                 colors = RColorBrewer::brewer.pal(n = 9, name = "Reds"), 
                 label_by = "sample_id", 
                 slot = "counts",
                 pt_size = 0.8, 
                 add_scalebar = TRUE,
                 scalebar_gg = scalebar(x = 1000),
                 scalebar_position = c(0.64, 0.08), 
                 scalebar_height = 0.07,
                 scale_alpha = FALSE) &  
        labs(fill="log UMI count per spot") &
        theme(legend.position = "right", 
        legend.text = element_text(angle = 0),
        legend.margin = margin(t = 50, r = 0, b = 100, l = 0))
  plot_list[[gene]] <- plt
  ggsave(file = paste0("/vacuolar/log/", gene, "_STm_log_UMI_counts_BA.pdf"), plot = plt, dpi = 300, width = 10, height =8)
  
}

combined_plot <- wrap_plots(plot_list)
combined_plot
ggsave(file = paste0("/vacuolar/log/all_vacuolar_STm_log_UMI_counts_BA.pdf"), plot = combined_plot, dpi = 300, width = 30, height =20)
```

```{r}
#log housekeeping
plot_list <- list()
for (gene in housekeeping) {
  plt <- MapFeatures(se_salmo_only_log, #se_salmo_only_log or se_salmo_only
                 features = gene, 
                 #max_cutoff = 0.99, 
                 override_plot_dims = F, 
                 section_number = 1,
                 colors = RColorBrewer::brewer.pal(n = 9, name = "Reds"), 
                 label_by = "sample_id", 
                 slot = "counts",
                 pt_size = 0.8, 
                 add_scalebar = TRUE,
                 scalebar_gg = scalebar(x = 1000),
                 scalebar_position = c(0.64, 0.08), 
                 scalebar_height = 0.07,
                 scale_alpha = FALSE) &  
        labs(fill="log UMI count per spot") &
        theme(legend.position = "right", 
        legend.text = element_text(angle = 0),
        legend.margin = margin(t = 50, r = 0, b = 100, l = 0))
  plot_list[[gene]] <- plt
  ggsave(file = paste0("/housekeeping/log/", gene, "_STm_log_UMI_counts_BA.pdf"), plot = plt, dpi = 300, width = 10, height =8)
  
}

combined_plot <- wrap_plots(plot_list)
combined_plot
ggsave(file = paste0("/housekeeping/log/housekeeping_STm_log_UMI_counts_BA.pdf"), plot = combined_plot, dpi = 300, width = 30, height =20)
```

Get dataframe of Salmonella genes UMI counts per spot
```{r}
present_genes <- salmo_genes[salmo_genes %in% rownames(se_filtered_BA)]

cat("Found genes:", present_genes, "\n")

expr_raw <- FetchData(se_filtered_BA, vars = present_genes, layer = "counts")
expr_raw # df [10,969 × 19] = spots x Salmo genes

```

```{r}
# Sum across bacterial genes per spot
bact_sum <- rowSums(expr_raw > 0)  # number of bacterial genes detected per spot
bact_sum
# Define a threshold — at least 1 bacterial gene expressed
bact_positive_spots <- sum(bact_sum > 0)
bact_positive_spots # 999

# Total number of spots in the sample
total_spots <- ncol(se_filtered_BA)

cat("Bacterial-positive spots:", bact_positive_spots, "\n")
cat("Total spots:", total_spots, "\n")
```


```{r}
# Fraction (percentage)
cat("Fraction of bacterial-positive spots:", round(bact_positive_spots / total_spots * 100, 2), "%\n")
```

Fraction of bacterial-positive spots: 4.76 %

### Average number of S.Tm genes per spot
To understand how many of the bacterial genes have nonzero expression in each spot, averaged across all spots.
```{r}
# Count how many bacterial genes detected per spot
bact_sum <- rowSums(expr_raw > 0)

# Average across all spots
avg_bact_genes_per_spot <- mean(bact_sum)
avg_bact_genes_per_spot
```
0.1319172 average salmo genes per spot

### Avg number of salmo UMIs per spot
Sum all bacterial UMIs per spot, then average across spots.
```{r}
# Sum bacterial UMIs per spot
total_bact_UMI_per_spot <- rowSums(expr_raw)

# Average across all spots
avg_bact_UMI_per_spot <- mean(total_bact_UMI_per_spot)
avg_bact_UMI_per_spot

# Average across only among bacterial-positive spots (nonzero total) 
avg_bact_UMI_per_positive_spot <- mean(total_bact_UMI_per_spot[total_bact_UMI_per_spot > 0])
avg_bact_UMI_per_positive_spot
```
0.1707539 average bacterial UMIs per spot
1.874875 average bacterial UMIs per spot (in salmo-positive spots)

### Avg number of UMIs per spot for each salmo gene
“For each bacterial gene, what is the average number of UMIs per spot?”
```{r}
# Average UMI per spot for each bacterial gene (across all spots)
avg_UMI_per_gene <- colMeans(expr_raw)
avg_UMI_per_gene

#only spots where the gene is expressed
avg_UMI_per_gene_detected <- apply(expr_raw, 2, function(x) mean(x[x > 0]))
avg_UMI_per_gene_detected
```

```{r}
salmo_summary <- data.frame(
  Gene = colnames(expr_raw),
  Avg_UMI_all_spots = colMeans(expr_raw),
  Avg_UMI_detected_spots = apply(expr_raw, 2, function(x) mean(x[x > 0])),
  Detection_rate = colMeans(expr_raw > 0)
)
salmo_summary
```

```{r}
write.xlsx(salmo_summary, file = "/path/UMIs_per_spot_for_each_STm_gene_BA.xlsx", colNames=TRUE, rowNames=TRUE)
```


## 2. Whether multiple bacterial genes are co-expressed within a single spot, or whether only one (or a subset) is typically detected.

```{r}
expr_raw_gene_names <- expr_raw
#with gene names
gene_names <- read_excel("table_STm_geneIDs_and_gene_names.xlsx")

new_colnames <- gene_names$STm_gene_name
colnames(expr_raw_gene_names) <- new_colnames
```

Count how many genes are detected per spot (singletons)
```{r}
# Binary detection: 1 = detected, 0 = not detected
binary_expr <- expr_raw_gene_names > 0
binary_expr
# Number of bacterial genes detected per spot
num_genes_per_spot <- rowSums(binary_expr)

# Singleton spots (exactly one gene detected)
singleton_spots <- num_genes_per_spot == 1
table(singleton_spots)

#which gene
singleton_gene_per_spot <- apply(binary_expr[singleton_spots, ], 1, function(x) names(x)[which(x)])
head(singleton_gene_per_spot)
```

For each gene, how many singletons
```{r}
# Subset singleton spots
singleton_matrix <- binary_expr[singleton_spots, ]

# Identify the singleton gene in each spot
singleton_genes <- apply(singleton_matrix, 1, function(x) names(x)[which(x)])

# Count number of singleton spots per gene
singleton_counts <- table(singleton_genes)
singleton_counts
```

```{r}
library(ggplot2)

singleton_df <- as.data.frame(singleton_counts)
colnames(singleton_df) <- c("Gene", "SingletonSpots")

singleton_plt <- ggplot(singleton_df, aes(x = Gene, y = SingletonSpots)) +
  geom_bar(stat = "identity", alpha = 1, fill = "steelblue") +
  theme_minimal() +
  geom_text(aes(label=SingletonSpots), vjust=-0.5, color="black", size=3) +
  labs(x="S.Tm gene", y="Number of singleton spots", title = " Singleton spots per bacterial gene") +
  theme(axis.text.y = element_text(size = 6, hjust = 1, color = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),  # Remove panel background
        axis.ticks.length = unit(0.25, "cm"),  
        axis.text.x = element_text(size = 10, color = "black", angle = 0, hjust = 0.5))

singleton_plt

ggsave(file = paste0("singleton_spots_per_bacterial_gene.pdf"), plot = singleton_plt, dpi = 300, width = 10, height = 12)
```

Colocalisation (co-expression) between gene pairs
How often each pair of bacterial genes is detected in the same spot
```{r}
library(reshape2)

# Convert binary detection to numeric (0/1)
binary_expr_num <- binary_expr * 1
binary_expr_num
# Co-detection matrix: number of spots where each pair is co-expressed
coexpr_matrix <- t(binary_expr_num) %*% binary_expr_num
coexpr_matrix

#Fraction of spots co-expressing each pair
coexpr_fraction <- coexpr_matrix / nrow(expr_raw)
coexpr_fraction
```

Fisher’s exact test to see which gene pairs significantly colocalise
```{r}
STm_gene_name = c("sipC", "uhpT", "iroN", "grxA", "sitA", "yjjZ", "fruB", "prgJ", "ssaG", "zinT", "sseC", "pipB", "mgtC", "sifA", "znuA", "phoN", "infC", "groES", "lpp")

gene_pairs <- combn(STm_gene_name, 2, simplify = FALSE)
pvals <- sapply(gene_pairs, function(pair) {
  g1 <- pair[1]
  g2 <- pair[2]
  
  # Contingency table
  tab <- matrix(c(
    sum(binary_expr[,g1] & binary_expr[,g2]),   # both
    sum(binary_expr[,g1] & !binary_expr[,g2]),  # g1 only
    sum(!binary_expr[,g1] & binary_expr[,g2]),  # g2 only
    sum(!binary_expr[,g1] & !binary_expr[,g2])  # neither
  ), nrow=2)
  
  fisher.test(tab, alternative = "greater")$p.value
})

# Combine results
coexpr_results <- data.frame(
  gene1 = sapply(gene_pairs, `[`, 1),
  gene2 = sapply(gene_pairs, `[`, 2),
  pvalue = pvals,
  co_detected_spots = sapply(gene_pairs, function(pair) sum(binary_expr[,pair[1]] & binary_expr[,pair[2]]))
)

coexpr_results <- coexpr_results[order(coexpr_results$pvalue), ]
head(coexpr_results)

```
Visualize p values from fisher's test
```{r}
coexpr_results$negLogP <- -log10(coexpr_results$pvalue)

# Combine gene pairs into a single label
coexpr_results$Pair <- paste(coexpr_results$gene1, coexpr_results$gene2, sep = "-")

coexpr_plot <- ggplot(coexpr_results, aes(x = pvalue, y = Pair)) +
  geom_bar(stat = "identity", fill = "coral") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "black", linewidth = 0.8) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Gene pair") +
  xlab("p-value") + #-log10(p-value)
  ggtitle("Significant co-localizing bacterial gene pairs")

coexpr_plot
```

```{r}
ggsave(file = paste0("coexpressed_gene_pairs_by_significance.pdf"), plot = coexpr_plot, dpi = 300, width = 10, height = 20)
```



## 3.Whether lifestype-specific markers co-occur within the same spots.
Colocalisation with 2 categories
Correlation metric: Fisher’s exact test

```{r}
cyt <- c("SL1344-2863", "SL1344-3753", "SL1344-2762", "SL1344-0848", "SL1344-2841", "SL1344-4483", "SL1344-2183", "SL1344-2852")

cyt_names <- c("sipC", "uhpT", "iroN", "grxA", "sitA", "yjjZ", "fruB", "prgJ")

vac <- c("SL1344-1340", "SL1344-1198", "SL1344-1334", "SL1344-1027", "SL1344-3729", "SL1344-1161", "SL1344-1825", "SL1344-4255")
vac_names <- c("ssaG", "zinT", "sseC", "pipB", "mgtC", "sifA", "znuA", "phoN")
```

```{r}
binary_expr_names <- expr_raw_gene_names > 0  # TRUE/FALSE for each spot

cyt_expr <- binary_expr_names[, cyt_names]
vac_expr <- binary_expr_names[, vac_names]

# co-detection matrix
coexpr_matrix_cyt <- t(cyt_expr) %*% cyt_expr
coexpr_matrix_vac <- t(vac_expr) %*% vac_expr
```

Gene pair colocalisation within cytsolic 
```{r}
gene_pairs <- combn(cyt_names, 2, simplify = FALSE)

fisher_results <- lapply(gene_pairs, function(pair) {
  g1 <- pair[1]
  g2 <- pair[2]
  
  # contingency table
  tab <- matrix(c(
    sum(cyt_expr[,g1] & cyt_expr[,g2]),   # both
    sum(cyt_expr[,g1] & !cyt_expr[,g2]),  # g1 only
    sum(!cyt_expr[,g1] & cyt_expr[,g2]),  # g2 only
    sum(!cyt_expr[,g1] & !cyt_expr[,g2])  # neither
  ), nrow = 2)
  
  # Fisher test
  p <- fisher.test(tab, alternative = "greater")$p.value
  
  data.frame(gene1 = g1, gene2 = g2, co_detected_spots = tab[1], pvalue = p)
})

fisher_results <- bind_rows(fisher_results)

```

```{r}
fisher_results$Pair <- paste(fisher_results$gene1, fisher_results$gene2, sep = "-")

cyto_gene_pair <- ggplot(fisher_results, aes(x = pvalue, y = reorder(Pair, pvalue))) +
  geom_bar(stat = "identity", fill = "coral") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "blue", size = 1) +
  theme_bw() +
  xlab("P-value") +
  ylab("Gene pair") +
  ggtitle("Co-localization of cytosolic genes")

```

```{r}
ggsave(file = paste0("coexpressed_cytosolic_gene_pairs.pdf"), plot = cyto_gene_pair, dpi = 300, width = 10, height = 20)
```

Gene pair colocalisation within vacuolar 
```{r}
gene_pairs_vac <- combn(vac_names, 2, simplify = FALSE)

fisher_results_vac <- lapply(gene_pairs_vac, function(pair) {
  g1 <- pair[1]
  g2 <- pair[2]
  
  # contingency table
  tab <- matrix(c(
    sum(vac_expr[,g1] & vac_expr[,g2]),   # both
    sum(vac_expr[,g1] & !vac_expr[,g2]),  # g1 only
    sum(!vac_expr[,g1] & vac_expr[,g2]),  # g2 only
    sum(!vac_expr[,g1] & !vac_expr[,g2])  # neither
  ), nrow = 2)
  
  # Fisher test
  p <- fisher.test(tab, alternative = "greater")$p.value
  
  data.frame(gene1 = g1, gene2 = g2, co_detected_spots = tab[1], pvalue = p)
})

fisher_results_vac <- bind_rows(fisher_results_vac)

```

```{r}
fisher_results_vac$Pair <- paste(fisher_results_vac$gene1, fisher_results_vac$gene2, sep = "-")

vac_gene_pair <- ggplot(fisher_results_vac, aes(x = pvalue, y = reorder(Pair, pvalue))) +
  geom_bar(stat = "identity", fill = "coral") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "blue", size = 1) +
  theme_bw() +
  xlab("P-value") +
  ylab("Gene pair") +
  ggtitle("Co-localization of vacuolar genes")

```

```{r}
ggsave(file = paste0("coexpressed_vacuolar_gene_pairs.pdf"), plot = vac_gene_pair, dpi = 300, width = 10, height = 20)
```

### Permutation test for cytosolic or vacuolar genes
```{r}
# Observed: number of spots with all genes detected
obs_count <- sum(vac_genes_per_spot == 1)

# Permutation: randomly shuffle gene detection across spots many times
set.seed(123)
n_perm <- 1000
perm_counts <- replicate(n_perm, {
  shuffled <- apply(vac_expr, 2, sample)
  sum(rowSums(shuffled) == ncol(shuffled))
})

# Empirical p-value
p_value <- mean(perm_counts >= obs_count)
p_value

```
Permutation shows that, in general, the cytosolic genes do not co-occur in same spots. Same for vacuolar genes. (p_value = 1)


## 4. The distribution of the bacterial UMIs per spot (from Figure 2B it appears the maximum is ~10).

Show the overall distribution of Salmo+ spots in the 3 conditions (spatially) and show a histogram. 

```{r}
# Total bacterial UMIs per spot
total_bact_UMIs <- rowSums(expr_raw_gene_names)
summary(total_bact_UMIs)

df <- data.frame(
  spot = names(total_bact_UMIs),
  total_bact_UMI  = as.numeric(total_bact_UMIs)
)

MapFeaturesSummary(se_filtered_BA, features = "salmo_counts", section_number = 2, subplot_type = "histogram") #up to 27 UMIs
```

Do not visualise with violin plot as a lot of 0 values

### Histogram of number of spots according to UMI counts
Only spots with UMIs > 0 are shown
```{r}
#Histogram or box plot of the distribution
umis_per_spot <- ggplot(df %>% filter(total_bact_UMI > 0), aes(x = total_bact_UMI)) +
  geom_histogram(binwidth = 1,  fill = "skyblue",
    color = "black") +
  scale_x_continuous(breaks = seq(min(df$total_bact_UMI), max(df$total_bact_UMI), by = 1)) +
  labs(title = "Distribution of the S.Tm UMIs per spot",
       x = "S.Tm raw UMI count",
       y = "Number of spots") +
  theme(axis.text.y = element_text(hjust = 1, color = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),  # Remove panel background
        axis.ticks.length = unit(0.25, "cm"),  # Adjust the length of the tick marks if needed
        axis.text.x = element_text(color = "black")) 

umis_per_spot
```

```{r}
ggsave(file = paste0("STm_umis_per_spot_v2.pdf"), plot = umis_per_spot, dpi = 300, width = 10, height = 6)
```



```{r}
df$is_zero <- df$total_bact_UMI == 0

spots_with_without_stm <- ggplot(df, aes(x = is_zero)) +
  geom_bar(width = 0.5, fill = "skyblue",
    color = "black") +
  scale_x_discrete(labels = c("S.Tm UMIs present", "S.Tm UMIs absent")) +
  labs(x = "S. Tm UMI value", y = "Number of spots") +
  theme(axis.text.y = element_text(hjust = 1, color = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),  # Remove panel background
        axis.ticks.length = unit(0.25, "cm"),  # Adjust the length of the tick marks if needed
        axis.text.x = element_text(color = "black"))

spots_with_without_stm
```

```{r}
ggsave(file = paste0("spot_count_with_vs_without_STm.pdf"), plot = spots_with_without_stm, dpi = 300, width = 10, height = 6)
```


```{r}
df_no_zero <- df %>% filter(total_bact_UMI > 0)
df_hist <- df_no_zero %>%
  count(total_bact_UMI)

ggplot(df_hist, aes(x = n, y = factor(total_bact_UMI))) +
  geom_col(fill = "skyblue", color = "black", width = 0.8) +
  labs(
    x = "Number of spots (cells)",
    y = "UMI count"
  ) +
  theme_minimal()
```

# Session info

<details>

<summary>

</summary>

```{r}
sessionInfo()
Sys.Date()
Sys.time()
```
